Paper 3: Housing and the capital share of income

### 2023-04-28 PhD_pages/Paper_1/paper1.qmd
Paper 1: Drafting

Their use and impact, however, has taken place within largely separate arenas. Many methodological differences separate the two datasets, including the  underlying data sources they rely on, the methods and assumptions applied to those sources, the welfare concept, the summary indices of inequality favoured, and coverage they provide. This lack of alignment between the datasets greatly limits our ability to combine the insights provided by each into a coherent empirical view informed by both. The two datasets represent distinct and, to a large extent, incommensurable 'worlds' of inequality measurement.

Moreover, these two worlds of measurement yield different views concerning the nature and evolution of inequality.

Firstly, while the *WID* data and its associated global reports suggest rising inequality was a near-ubiquitous phenomenon [@alvaredo_world_2018; @chancel_world_2021], in contrast the World Bank *PIP* data and its associated reports offer a considerably more mixed picture, in which inequality has been rising and falling in roughly equal numbers of countries [@world_bank_poverty_2016].[^WID_PIP_quotes]

[^WID_PIP_quotes]:
    > Income and wealth inequalities have been on the rise nearly everywhere since the 1980s, following a series of deregulation and liberalization programs which took different forms in different countries. [@chancel_world_2021, pp. xx]

@@

    > [Since c.1993] the number of countries with rising inequality was **slightly higher** than the number of countries with falling inequality. Yet, the average Gini among these 91 countries actually **declined**. [@world_bank_poverty_2016, pp. xx]

Secondly, the datasets provide contrasting accounts as to how inequalities *within* countries compare to the differences in average incomes seen *between* countries. Most analyses of global income inequality to date have relied heavily on the *PIP* data [@Kanbur2022-lk; @lakner_global_2016; @anand_global_2015; @milanovic_global_2015; @Nino-Zarazua2017-tv; @Ravallion2014-mk]. A consistent result from these studies is that, despite recent convergence [@patel_new_2021], differences in average incomes between countries continue to make up the lion's share of global inequality. On such accounts, a person's income is largely determined by the country in which they happen to have been born [@milanovic_global_2015]. More recent analysis of the *WID* data, however, flips this established result on its head. @chancel_global_2021 find within-country inequalities instead 'dominate' inequality between countries, contributing 78% in a decompositon of the global Theil index.

Thus despite the greatly improved range of distributional data at our disposal, considerable ambiguity continues to surround some of the most basic descriptive questions concerning inequality around the world.

@@

# Bridging the gap for particular countries

>>>>

Their use and impact, however, has taken place within largely separate arenas. Many methodological differences separate the two datasets, including the  underlying data sources they rely on, the methods and assumptions applied to those sources, the concept of welfare measured, the summary indices of inequality typically favoured, and coverage they provide. The two datasets represent distinct and, to a large extent, incommensurable 'worlds' of inequality measurement. This lack of alignment between the datasets limits our ability to combine the insights provided by each into a coherent empirical view informed by both.

This disconnect is particularly problematic given the starkly contrasting perspectives on inequality the two worlds provide.

Analysis of the *PIP* data suggests that within-country inequality trends have been decidedly mixed in recent decades. Inequality is found to have fallen in roughly as many countries as it has risen, with no substantial trend in either direction on average [@world_bank_poverty_2016].[^PIP_quotes] In contrast, summaries of the *WID* data emphasise that rising inequality within countries has been an almost universal phenomenon [@alvaredo_world_2018; @chancel_world_2021].[^WID_quotes]

[^PIP_quotes]:
    > [Since c.1993] the number of countries with rising inequality was **slightly higher** than the number of countries with falling inequality. Yet, the average Gini among these 91 countries actually **declined**. [@world_bank_poverty_2016, pp. xx]

@@

The two datasets have also yielded starkly contrasting accounts of how big a contribution within-country inequalities make to overall global inequality. Most analyses of global income inequality to date have relied heavily on the *PIP* data [@kanbur_global_2022; @lakner_global_2016; @anand_global_2015; @milanovic_global_2015; @Nino-Zarazua2017-tv]. A consistent result from these studies is that, despite some recent convergence in average incomes between countries [@patel_new_2021], such differences  continue to make up the lion's share of global inequality. On such accounts, a person's income is largely determined purely by the country in which they happen to have been born [@milanovic_global_2015]. More recent analysis of the *WID* data, however, flips this established result on its head. @chancel_global_2021 find within-country inequalities instead 'dominate' inequality between countries, contributing 78% in a decompositon of the global Theil index.

The two worlds of *measurement* thus depict two very different empirical *realities*. Knowing which of these worlds we live in matters not just for our understanding of inequality today, but also our understanding of how we should seek to combat inequality in the future. A near ubiquitous rise in inequality apparent in the *WID* data points to fundamental 'forces for divergence' within capitalism [@piketty_capital_2014] and the need for global coordination, particularly on the taxation of capital, to counter them [@saez_triumph_2019]. If inequalities within countries now predominate within the global distribution of income, such avenues will also be central to tackling global inequality too [@piketty_brief_2022]. Conversely, the much more mixed picture seen in the *PIP* data highlights more prominently the past role and future scope of national policies to shape the extent of inequality within countries [@ravallion_inequality_2018]. From this viewpoint, such inequalities in any case remain a small share of overall global inequality such that the global distribution will continue to depend heavily on future aggregate growth rates [@kanbur_global_2022].

More fundamentally, the two datasets support differing narratives concerning globalization. [TODO: elabourate].

Despite laudable efforts to make transparent, in practice a black box. Treatmetn of the underlying data and assumptions. Whilst strengths and weaknesses studied for particular countries with good data, when applied globally we really don't know. Despite the great data, fundamental ambiguity surrounding the most basic aspects.

@@

# Bridging the gap for particular countries / welfare concept

"Korinek, Mistiaen, and
Ravallion (2006) estimate that correcting
for such selective compliance adds about
five percentage points to the Gini index for
the United States."

### 2023-04-28 PhD_pages/older_structure/paper_pages/paper3.qmd
Restructuring folders so that everything is contained in one complete subfolder per paper
Paper 1: Drafting

>>>>

NB: I should look at the global estimates of factor shares mentioned in some UN reports (maybe ILO data?). e.g. 
- "The share of wages in total GDP declined in a majority of countries
(91 out of 133 with data) from 1995 to 2014 (ILO, 2016)." (https://www.un.org/development/desa/dspd/wp-content/uploads/sites/22/2020/02/World-Social-Report2020-FullReport.pdf).
- "A decline in the labour share of income from 2014 to 2019 ‐ from 54.1 per cent to 52.6 per cent ‐ represents upward pressure on inequality." https://www.un.org/sustainabledevelopment/inequality/
